A Plasma Torus Around a Young Low-Mass Star

### 2025-04-07 paper/ms.tex
update ms

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time the existence of corotating clumps of circumstellar plasma around a CPV.
Given that cool ($10^4$ K) plasma can persist in the hot ($10^6$ K)
coronae of rapidly-rotating stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that the magnetospheres of the
lowest-mass stars can support long-lived, optically thick
clumps of material that explain the sharp photometric features seen in
CPVs.  The origin of this optically thick material and its
microphysical opacity remain unclear.
\end{abstract}

@@

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and peak velocities, this material's motion
is not controlled by gravitational attraction to the star; it is more
readily explained as a partially-ionized plasma being dragged along with
a rigidly rotating stellar magnetic field.  The velocity semi-amplitude
of the sinusoids gives the mean distance of each clump from the stellar
surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the inner clumps, and
3.88\,$R_\star$ for the outer clump.   These clumps transit in front of
the star when passing from negative to positive velocity.  The innermost
clump transits over $\approx$15\% of the period, at $\phi$=0 and
$\phi$=1.  This spectroscopic transit happens during the latter half of
the complex photometric feature in the TESS data.

@@

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they show for the first time that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth photometric
variability signals, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} seems to be incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission coming from a region half the size of the
star (see \S~\ref{subsec:halpha}).  The observed H$\alpha$ luminosity
suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see \S~\ref{subsec:gas}).  Dust, if present, is
independently constrained to have a total mass $10^{15}\,{\rm g} <
M_{\rm dust} < 10^{17}\,{\rm g}$ (see \S~\ref{subsec:dust}).  This
circumstellar material could originate either from the star or from an
external source.  Plausible external sources include an undetected old
disk, comets, or a close-in exoplanet.  This latter scenario would
make CPVs extrasolar analogs of the Jupiter-Io plasma torus
(e.g.~Ref.~\cite{Bagenal1981}).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES
  observation (green bar). 
  {\bf a,} TESS simple aperture photometry.  The main data gaps were
  caused by scattered light from the Earth (BTJD 3356-3358.5) and Moon
  (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.
  {\bf b-c,} Folded light curve before and after spectroscopy.  Raw
  two minute data are in gray; black phase-averages to 100 points per
  cycle.  During BTJD 3352-3356, a state switch occurred near BTJD
  3353, in which the dip at $\phi$$\approx$0.8 disappeared.  While the
  large eclipse feature was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.
  }
  \label{fig:fulllc}
\end{figure}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf7.pdf}
  \caption{Orbits fit to mean radial velocities (RVs) extracted from 
  H$\alpha$ profile fits in Figure~\ref{fig:halphamodel}.
  Radial velocities on the vertical axis are in units of the
  equatorial velocity, $v_{\rm eq}$=130\,\kms.
  Each marker denotes the best-fit mean at a given epoch.
  Solid circles were adopted in the fits; transparent X markers
  were excluded due to concerns about their reliability
  (see text).  The inner double-peaked clump is shown in
  black and orange; the outer clump is shown in green.  Five draws from
  each posterior probability distribution are shown.}
  \label{fig:orbits}
\end{figure}

Accounting for these caveats, we used the results from Extended Data
Table~\ref{tab:halphamodelparams} to quantify the orbital periods and
velocity semi-amplitudes of the clumps.
Figure~\ref{fig:orbits} shows the results assuming circular
orbits; considering the Bayesian information criterion, we found no
reason to prefer eccentric orbits.  By visually
inspecting Extended Data Figure~\ref{fig:halphamodel}, we excluded
epochs where our gaussian profile fitting failed to either detect or
else adequately represent the circumstellar emission.  We then used
NumPyro to sample the Gaussian likelihood for a Keplerian orbit
with the NUTS algorithm \cite{Phan2019}.  We used the measurement
uncertainties from each estimated mean radial velocity value and
included an additional jitter term in quadrature.  This procedure
yielded orbital periods and semi-amplitudes of
$P_0$=$3.92\pm0.03$\,hr, $K_0/v_{\rm eq}$=2.07$\pm$0.04;
$P_1$=$3.92\pm0.06$\,hr, $K_1/v_{\rm eq}$=2.88$\pm$0.10;
and $P_2$=$3.88\pm0.20$\,hr, $K_1/v_{\rm eq}$=3.88$\pm$0.25.
The periods for the inner double-peaked clump are therefore consistent with the
photometric $3.930\pm0.001$\,hr period within a precision of two
minutes.  The ``period'' for the outer clump is ambiguous because the
H$\alpha$ data only support the idea of a periodic orbit of material
well-fit by gaussian emission from $\phi$$\approx$0.5-1.0.  From
$\phi$=0-0.2, there is no detectable emission, and from
$\phi$=0.2-0.4, the emission spans 1-4\,$v_{\rm eq}$ without taking a
clear gaussian shape.  While the outer-most edge of this
$\phi$=0.2-0.4 emission provides a plausible match to the expectation
of a circular orbit, the idea of invoking a particular functional form
for this component seems fine-tuned.  We instead emphasize that
although this emission is present, its variability in time is 
inconsistent with the idea of a stable clump of material.
Additional observations would be needed to conclusively
determine whether or not this component of the system is long-lived.

@@

Circumstellar H$\alpha$ emission might be sourced either from resonant
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figure~\ref{fig:spec}
and Extended Data Figure~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination can be written
\begin{equation} j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm
eff}_{\rm H\alpha} h \nu_{\rm H\alpha}, \end{equation} where $n_{\rm
e}$ and $n_{\rm p}$ are the electron and proton densities, and
$\alpha^{\rm eff}_{\rm H\alpha}$ is the effective recombination
coefficient, defined to include all recombination routes that produce
an H$\alpha$ photon.  For hydrogen with temperatures between
1,000-10,000\,K, $\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the
order of $10^{-12}$ to $10^{-13}$\,cm$^3$\,s$^{-1}$
\cite{Hummer1987,Draine2011}.  Neglecting the effects of atoms other
than hydrogen, we can assume an ionization fraction $x$, such that
$n_{\rm e} = n_{\rm p} = x n_{\rm H}$, for $n_{\rm H}$ the hydrogen
number density.  Let $L_{\rm H\alpha} = j_{\rm H\alpha} V$, for $V$
the volume of the emitting hydrogen.  The luminosity of circumstellar
hydrogen emission, $L_{\rm H\alpha}$, is an observable: our SED
fitting routine yields $L_\star$$\approx$0.012\,$L_\odot$, which
implies that the stellar H$\alpha$ line radiates at
$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$.  The luminosity of the
clumps $L_{\rm H\alpha}$ are of order one tenth that of the star.  If
we approximate the emitting volume as a homogeneous sphere of radius
$r$, we can write
\begin{align}
  n_{\rm H} &= 
  1 \cdot 10^{11}\,{\rm cm}^{-3}
  \left(
    \frac{0.5}{x}
  \right)
  \left( 
    \frac{ L_{\rm H\alpha} }{ 10^{27}\,{\rm erg\,s^{-1}} }
    \frac{ 10^{-13}\,{\rm cm^{3}\,s^{-1}} }{ \alpha^{\rm eff}_{\rm H\alpha} }
  \right)^{1/2}
  \left(
    \frac{ 0.1\,R_\odot }{ r }
  \right)^{3/2}.
  \label{eq:numdensity}
\end{align}
For a uniform density clump, this suggests a total gas mass of $M_{\rm
gas}\approx 8\times10^{16}$\,g.
We emphasize that Equation~\ref{eq:numdensity} in intended to provide
only an order of magnitude estimate for the number density implied by
the observed H$\alpha$ emission.  In detail, the effective
recombination rate and the ionization fraction each vary with density
and temperature; a more thorough estimate would iteratively solve the
equations of detailed balance and radiative transfer
(e.g.~\cite{CollierCameron1989} Figure~8), and potentially also
consider departures from local thermodynamic equilibrium.

@@

To summarize, if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$8$\times$10$^{16}$\,g from
\S~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an
accumulation of externally-sourced material.  While there are several
plausible external sources, feeding through a low-mass disk in
particular cannot be ruled out based on typical disk depletion times
\cite{Haisch2001}.  Observations of infrared excesses and accretion
signatures in low-mass stars tens of millions of years old suggest a
broad lifetime distribution for such disks
\cite{Silverberg2020,Lee2020,Gaidos2022,Pfalzner2024}.

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time that corotating clumps of circumstellar plasma exist around a CPV.
Given that cool ($10^4$ K) plasma can persist in the hot ($10^6$ K)
coronae of rapidly-rotating stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that the magnetospheres of red dwarfs can
support long-lived, optically thick clumps of material that produce
the sharp photometric features seen in CPVs.  The origin of this
optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and velocities, this material's motion
is not controlled by gravitational attraction to the star; it is more
readily explained as a partially-ionized plasma being dragged along with
a rigidly rotating stellar magnetic field.  The velocity semi-amplitude
of the sinusoids gives the mean distance of each clump from the stellar
surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the inner clumps, and
3.88\,$R_\star$ for the outer clump.   These clumps transit in front of
the star when passing from negative to positive velocity.  The
transits of the two inner clumps last $\approx$22\% of each cycle,
from $\phi$=-0.1 to $\phi$=+0.12 (see \S~\ref{subsec:halpha}).  This
spectroscopic transit coincides with the latter half of the complex
eclipse feature in the TESS data.

@@

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they show for the first time that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth signals to the photometric
variability, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} seems to be incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission in each clump coming from a region half
the size of the star (see \S~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see \S~\ref{subsec:gas}).  Dust, if present, is
independently constrained to have a total mass $10^{15}\,{\rm g} <
M_{\rm dust} < 10^{17}\,{\rm g}$ (see \S~\ref{subsec:dust}).  This
circumstellar material could originate either from the star or from an
external source.  Plausible external sources include an undetected old
disk, comets, or a close-in exoplanet.  This latter scenario would
make CPVs extrasolar analogs of the Jupiter-Io plasma torus
(e.g.~Ref.~\cite{Bagenal1981}).

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the Keck/HIRES
  observation (green bar). 
  {\bf a,} TESS simple aperture photometry.  The main data gaps were
  caused by scattered light from the Earth (BTJD 3356-3358.5) and Moon
  (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.
  {\bf b-c,} Folded TESS light curve before and after spectroscopy.  Raw
  two minute data are in gray; black phase-averages to 100 points per
  cycle.  During BTJD 3352-3356, a state switch occurred near BTJD
  3353, in which the dip at $\phi$$\approx$0.8 disappeared.  While the
  large eclipse feature was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.
  The oranges bar denote times of spectroscopic transits for the inner
  two H$\alpha$ clumps observed with HIRES.
  }
  \label{fig:fulllc}
\end{figure}

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf7.pdf}
  \caption{{\bf a,} Orbits fit to mean radial velocities (RVs)
  extracted from H$\alpha$ profile fits in
  Figure~\ref{fig:halphamodel}.  Radial velocities on the vertical
  axis are in units of the equatorial velocity, $v_{\rm
  eq}$=130\,\kms.  Each marker denotes the best-fit gaussian mean at a
  given epoch.  Solid circles were adopted in the fits; transparent X
  markers were excluded due to reliability concerns (see text).  The
  inner double-peaked clump is shown in violet and orange; the outer
  clump is shown in green.  Five model draws from each posterior probability
  distribution are plotted.  {\bf b,} Idealized emissivity from mean
  model fits in Panel a, assuming a constant emission amplitude and
  velocity width $\sigma$=0.25\,$v_{\rm eq}$ over the observation
  sequence.}
  \label{fig:orbits}
\end{figure}

Accounting for these caveats, we used the results from Extended Data
Table~\ref{tab:halphamodelparams} to quantify the orbital periods and
velocity semi-amplitudes of the clumps.
Figure~\ref{fig:orbits}a shows the results assuming circular
orbits; considering the Bayesian information criterion, we found no
reason to prefer eccentric orbits.  By visually
inspecting Extended Data Figure~\ref{fig:halphamodel}, we excluded
epochs where our gaussian profile fitting failed to either detect or
else adequately represent the circumstellar emission.  We then used
NumPyro to sample the Gaussian likelihood for a Keplerian orbit
with the NUTS algorithm \cite{Phan2019}.  We used the measurement
uncertainties from each estimated mean radial velocity value and
included an additional jitter term in quadrature.  This procedure
yielded orbital periods and semi-amplitudes of
$P_0$=$3.92\pm0.03$\,hr, $K_0/v_{\rm eq}$=2.07$\pm$0.04;
$P_1$=$3.92\pm0.06$\,hr, $K_1/v_{\rm eq}$=2.88$\pm$0.10;
and $P_2$=$3.88\pm0.20$\,hr, $K_1/v_{\rm eq}$=3.88$\pm$0.25.
The periods for the inner double-peaked clump are therefore consistent with the
photometric $3.930\pm0.001$\,hr period within a precision of two
minutes.  The ``period'' for the outer clump is ambiguous because the
H$\alpha$ data only support the idea of a periodic orbit of material
well-fit by gaussian emission from $\phi$$\approx$0.5-1.0.  From
$\phi$=0-0.2, there is no detectable emission, and from
$\phi$=0.2-0.4, the emission spans 1-4\,$v_{\rm eq}$ without taking a
clear gaussian shape.  While the outer-most edge of this
$\phi$=0.2-0.4 emission provides a plausible match to the expectation
of a circular orbit, the idea of invoking a particular functional form
for this component seems fine-tuned.  We instead emphasize that
although this emission is present, its variability in time is 
inconsistent with the idea of a stable clump of material.
Additional observations would be needed to conclusively
determine whether or not this component of the system is long-lived.

For Figure~\ref{fig:orbits}b, we then used the mean orbits
from Figure~\ref{fig:orbits}a to generate sinusoidal-in-time
gaussians, similar to the observations.  We assumed a constant
emission amplitude and velocity width $\sigma$=0.25\,$v_{\rm eq}$ for
this exercise, and normalized each gaussian to unit amplitude;  the
colorbar in Figure~\ref{fig:orbits}b thus masks the non-physical
additive contribution near the zero-crossing of velocity.
Compared to the behavior of the data at $\phi$=0.2-0.5
(Figure~\ref{fig:spec}), this is highly idealized.
Nonetheless, this exercise indicates that the transit of the inner
clump lasts $\approx$22\% of each cycle, with a slight asymmetry
around $\phi$=0.

@@

Circumstellar H$\alpha$ emission might be sourced either from resonant
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figure~\ref{fig:spec}
and Extended Data Figure~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination can be written
\begin{equation} j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm
eff}_{\rm H\alpha} h \nu_{\rm H\alpha}, \end{equation} where $n_{\rm
e}$ and $n_{\rm p}$ are the electron and proton densities, and
$\alpha^{\rm eff}_{\rm H\alpha}$ is the effective recombination
coefficient, defined to include all recombination routes that produce
an H$\alpha$ photon.  For hydrogen with temperatures between
1,000-10,000\,K, $\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the
order of $10^{-12}$ to $10^{-13}$\,cm$^3$\,s$^{-1}$
\cite{Hummer1987,Draine2011}.  Neglecting the effects of atoms other
than hydrogen, we can assume an ionization fraction $x$, such that
$n_{\rm e} = n_{\rm p} = x n_{\rm H}$, for $n_{\rm H}$ the hydrogen
number density.  Let $L_{\rm H\alpha} = j_{\rm H\alpha} V$, for $V$
the volume of the emitting hydrogen.  The luminosity of circumstellar
hydrogen emission, $L_{\rm H\alpha}$, is an observable: our SED
fitting routine yields $L_\star$$\approx$0.012\,$L_\odot$, which
implies that the stellar H$\alpha$ line radiates at
$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$.  The luminosity of the
clumps $L_{\rm H\alpha}$ are of order one tenth that of the star.  If
we approximate the emitting volume as a homogeneous sphere of radius
$r$, we can write
\begin{align}
  n_{\rm H} &= 
  1 \cdot 10^{11}\,{\rm cm}^{-3}
  \left(
    \frac{0.5}{x}
  \right)
  \left( 
    \frac{ L_{\rm H\alpha} }{ 10^{27}\,{\rm erg\,s^{-1}} }
    \frac{ 10^{-13}\,{\rm cm^{3}\,s^{-1}} }{ \alpha^{\rm eff}_{\rm H\alpha} }
  \right)^{1/2}
  \left(
    \frac{ 0.1\,R_\odot }{ r }
  \right)^{3/2}.
  \label{eq:numdensity}
\end{align}
For a uniform density clump, this suggests a total gas mass of $M_{\rm
gas}\approx 2\times10^{17}$\,g.
We emphasize that Equation~\ref{eq:numdensity} in intended to provide
only an order of magnitude estimate for the number density implied by
the observed H$\alpha$ emission.  In detail, the effective
recombination rate and the ionization fraction each vary with density
and temperature; a more thorough estimate would iteratively solve the
equations of detailed balance and radiative transfer
(e.g.~\cite{CollierCameron1989} Figure~8), and potentially also
consider departures from local thermodynamic equilibrium.

@@

To summarize, if dust is responsible for
the broadband variability of CPVs, it would need to be concentrated in
clumps with masses in the range of $10^{15}$-$10^{17}$\,g.  Given
$M_{\rm gas}$$\approx$2$\times$10$^{17}$\,g from
\S~\ref{subsec:gas}, the allowed dust masses imply $M_{\rm
gas}/M_{\rm dust}$ ranges of 1-100.  More careful measurements of this
ratio---in particular by inferring the dust mass through high
precision infrared spectrophotometry---could provide a path for
distinguishing the scenario of a trapped stellar outflow from an
accumulation of externally-sourced material.  While there are several
plausible external sources, feeding through a low-mass disk in
particular cannot be ruled out based on typical disk depletion times
\cite{Haisch2001}.  Observations of infrared excesses and accretion
signatures in low-mass stars tens of millions of years old suggest a
broad lifetime distribution for such disks
\cite{Silverberg2020,Lee2020,Gaidos2022,Pfalzner2024}.
